---
title: "Exercise 10"
author: "Sarah Lötscher"
date: "20 11 2022"
format:
  html:
    self-contained: true
---


```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


## Preprocessing and Clustering of a PBMC single cell RNA-seq data

The single cell dataset consists of peripheral blood mononuclear cells. These cells comprise different cell types. For each cell the ground truth is provided in terms of the assigned cell type that was derived using additional data. A cell has "unassigned" as cell type if it could not be reliably assigned to a cell type. 

1. Modify the quality-based cell filtering to remove the low quality cells that could not be assigned. Try with  different thresholds 
2. Try different clustering methods and parameters such that different cell types are in separate clusters.

Below is an example workflow inspired by http://bioconductor.org/books/release/OSCA/


```{r load packages, warning=FALSE}
library(stats)
suppressPackageStartupMessages(library(scran))
suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(BiocParallel))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(HDF5Array))
suppressPackageStartupMessages(library(hdf5r))
suppressPackageStartupMessages(library(pheatmap))
BPPARAM = BiocParallel::MulticoreParam(workers=2)
BPPARAM = BiocParallel::SerialParam()


```

#Default Kmeans clustering with modifyed quality-based cell filtering 

## Load the data

```{r setup nk, include=TRUE}
startTime = proc.time()
```

```{r load data}
sce = readRDS("pbmc-sce.RDS")
sce
```

Optionally subsample 2000 cells to reduce runtime
```{r subsample cells}
set.seed(38)
sce = sce[ , sample(1:ncol(sce), size = 2000, replace = FALSE)]

```

The cell type abundandances are
```{r barplot cell types, fig.width=10, fig.height=5}
cd = as.data.frame(colData(sce))
ggplot(cd, aes(x=cellType)) + geom_bar() +
   scale_fill_manual() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))


```

## Modify the quality-based cell filtering to remove the low quality cells that could not be assigned

Compute quality scores
```{r quality scores}

isMito <- grepl("^mt-", rowData(sce)$Symbol, ignore.case=TRUE)

sce <- addPerCellQC(sce, subsets=list(Mito=isMito), percent_top=100,
                    detection_limit=0, BPPARAM=BPPARAM)

```




Filter by quality
```{r filter low quality cells}
qc.lib <- isOutlier(sce$sum, log=TRUE, type="lower",nmads = 1.8)
qc.nexprs <- isOutlier(sce$detected, log=TRUE, type="lower",nmads = 1.8)
qc.mito <- isOutlier(sce$subsets_Mito_percent, type="higher",nmads = 1.8)
qc.top <- isOutlier(sce$percent.top_100, type="higher",nmads = 1.8)

discard <- qc.lib | qc.nexprs | qc.mito | qc.top

DataFrame(LibSize=sum(qc.lib), NExprs=sum(qc.nexprs),
          MitoProp=sum(qc.mito),
          TopProp=sum(qc.top), Total=sum(discard))

sce$discard = discard

sceFilt = sce[ , !sce$discard]
minCellsExpressed = 5
isExpressed <- Matrix::rowSums(counts(sceFilt) >= 1) >= minCellsExpressed
sceFilt = sceFilt[isExpressed, ]

```

Normalize

```{r normalize}
sceFilt = logNormCounts(sceFilt)


```

Compute reduced dimension representation
```{r dim red}
set.seed(38)
dec <- modelGeneVarByPoisson(sceFilt, BPPARAM=BPPARAM)
topGenes <- getTopHVGs(dec, n = 2000)
sceFilt <- runPCA(sceFilt, subset_row=topGenes, BPPARAM=BPPARAM) 
sceFilt <- runUMAP(sceFilt, dimred="PCA", BPPARAM=BPPARAM)

```

Cluster
```{r kmeans}
set.seed(100)
clust.kmeans <- kmeans(reducedDim(sceFilt, "PCA"), centers=10)
table(clust.kmeans$cluster)

```



```{r plot umap, fig.width=10}
colLabels(sceFilt) <- factor(clust.kmeans$cluster)
plotReducedDim(sceFilt, "UMAP", colour_by="label")
plotReducedDim(sceFilt, "UMAP", colour_by="cellType")
```

Agreement of clusters with cell types
```{r cluster agreement}
tab <- table(sceFilt$label, sceFilt$cellType)
phm <- pheatmap(log10(tab+10), main="cluster vs cell type",
    color=viridis::viridis(100), silent=FALSE)
phm


```




## Result

Sensitivity and specificity of low quality cell detection
```{r sensitivity}
mean(sce$discard[sce$cellType == "unassigned"])
mean(!sce$discard[sce$cellType != "unassigned"])

```

Agreement of clustering and cell type

```{r adjusted rand index}
mclust::adjustedRandIndex(sceFilt$label,sceFilt$cellType)
```

Runtime in seconds
```{r elapsed}
paste(signif((proc.time() - startTime)[ "elapsed"], digits=4), "s")
```




# 2) Graph free Clustering with different thresholds for quality-based cell filtering

```{r setup GFC, include=TRUE}
startTime = proc.time()
```

## Load the data

```{r load data GFC}
sce = readRDS("pbmc-sce.RDS")
sce
```

Optionally subsample 2000 cells to reduce runtime
```{r subsample cells GFC}
set.seed(38)
sce = sce[ , sample(1:ncol(sce), size = 2000, replace = FALSE)]

```

The cell type abundandances are
```{r barplot cell types GFC, fig.width=10, fig.height=5}
cd = as.data.frame(colData(sce))
ggplot(cd, aes(x=cellType)) + geom_bar() +
   scale_fill_manual() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))


```

## Modify the quality-based cell filtering to remove the low quality cells that could not be assigned

Compute quality scores
```{r quality scores GFC}

isMito <- grepl("^mt-", rowData(sce)$Symbol, ignore.case=TRUE)

sce <- addPerCellQC(sce, subsets=list(Mito=isMito), percent_top=100,
                    detection_limit=0, BPPARAM=BPPARAM)

```




Filter by quality
```{r filter low quality cells GFC}
qc.lib <- isOutlier(sce$sum, log=TRUE, type="both",nmads = 1.75)
qc.nexprs <- isOutlier(sce$detected, log=TRUE, type="both",nmads = 1.75)
qc.mito <- isOutlier(sce$subsets_Mito_percent, type="both",nmads = 1.75)
qc.top <- isOutlier(sce$percent.top_100, type="both",nmads = 1.75)

discard <- qc.lib | qc.nexprs | qc.mito | qc.top

DataFrame(LibSize=sum(qc.lib), NExprs=sum(qc.nexprs),
          MitoProp=sum(qc.mito),
          TopProp=sum(qc.top), Total=sum(discard))

sce$discard = discard

sceFilt = sce[ , !sce$discard]
minCellsExpressed = 5
isExpressed <- Matrix::rowSums(counts(sceFilt) >= 1) >= minCellsExpressed
sceFilt = sceFilt[isExpressed, ]

```

Normalize

```{r normalize GFC}
sceFilt = logNormCounts(sceFilt)


```

Compute reduced dimension representation
```{r dim red GFC}
set.seed(38)
dec <- modelGeneVarByPoisson(sceFilt, BPPARAM=BPPARAM)
topGenes <- getTopHVGs(dec, n = 2000)
sceFilt <- runPCA(sceFilt, subset_row=topGenes, BPPARAM=BPPARAM) 
sceFilt <- runUMAP(sceFilt, dimred="PCA", BPPARAM=BPPARAM)

```


Cluster
```{r GFC}
set.seed(100)
library(bluster)
nn.clusters <- clusterCells(sceFilt, use.dimred="PCA",BLUSPARAM= NNGraphParam(k=7))
table(nn.clusters)

```



```{r plot umap GFC, fig.width=10}
colLabels(sceFilt) <- nn.clusters
plotReducedDim(sceFilt, "UMAP", colour_by="label")
plotReducedDim(sceFilt, "UMAP", colour_by="cellType")
```

Agreement of clusters with cell types
```{r cluster agreement GFC}
tab <- table(sceFilt$label, sceFilt$cellType)
phm <- pheatmap(log10(tab+10), main="cluster vs cell type",
    color=viridis::viridis(100), silent=FALSE)
phm


```



## Result

Sensitivity and specificity of low quality cell detection
```{r sensitivity GFC}
mean(sce$discard[sce$cellType == "unassigned"])
mean(!sce$discard[sce$cellType != "unassigned"])

```

Agreement of clustering and cell type

```{r adjusted rand index GFC}
mclust::adjustedRandIndex(sceFilt$label,sceFilt$cellType)
```

Runtime in seconds
```{r elapsed GFC}
paste(signif((proc.time() - startTime)[ "elapsed"], digits=4), "s")
```



# 3) Hirarchical Clustering with different thresholds for quality-based cell filtering

```{r setup HC, include=TRUE}
startTime = proc.time()
```

## Load the data

```{r load data HC}
sce = readRDS("pbmc-sce.RDS")
sce
```

Optionally subsample 2000 cells to reduce runtime
```{r subsample cells HC}
set.seed(38)
sce = sce[ , sample(1:ncol(sce), size = 2000, replace = FALSE)]

```

The cell type abundandances are
```{r barplot cell types HC, fig.width=10, fig.height=5}
cd = as.data.frame(colData(sce))
ggplot(cd, aes(x=cellType)) + geom_bar() +
   scale_fill_manual() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))


```

## Modify the quality-based cell filtering to remove the low quality cells that could not be assigned

Compute quality scores
```{r quality scores HC}

isMito <- grepl("^mt-", rowData(sce)$Symbol, ignore.case=TRUE)

sce <- addPerCellQC(sce, subsets=list(Mito=isMito), percent_top=100,
                    detection_limit=0, BPPARAM=BPPARAM)

```




Filter by quality
```{r filter low quality cells HC}
qc.lib <- isOutlier(sce$sum, log=TRUE, type="both",)
qc.nexprs <- isOutlier(sce$detected, log=TRUE, type="both")
qc.mito <- isOutlier(sce$subsets_Mito_percent, type="both")
qc.top <- isOutlier(sce$percent.top_100, type="both")

discard <- qc.lib | qc.nexprs | qc.mito | qc.top

DataFrame(LibSize=sum(qc.lib), NExprs=sum(qc.nexprs),
          MitoProp=sum(qc.mito),
          TopProp=sum(qc.top), Total=sum(discard))

sce$discard = discard

sceFilt = sce[ , !sce$discard]
minCellsExpressed = 6 #Edited
isExpressed <- Matrix::rowSums(counts(sceFilt) >= 1) >= minCellsExpressed
sceFilt = sceFilt[isExpressed, ]

```

Normalize

```{r normalize HC}
sceFilt = logNormCounts(sceFilt)


```

Compute reduced dimension representation
```{r dim red HC}
set.seed(38)
dec <- modelGeneVarByPoisson(sceFilt, BPPARAM=BPPARAM)
topGenes <- getTopHVGs(dec, n = 2000)
sceFilt <- runPCA(sceFilt, subset_row=topGenes, BPPARAM=BPPARAM) 
sceFilt <- runUMAP(sceFilt, dimred="PCA", BPPARAM=BPPARAM)

```


```{r HC}
set.seed(100)
clust.hier <- hclust(dist(reducedDim(sceFilt, "PCA")), method="complete")
clust.hier <- cutree(clust.hier,k=15)
table(clust.hier)

```



```{r plot umap HC, fig.width=10}
colLabels(sceFilt) <- factor(clust.hier)
plotReducedDim(sceFilt, "UMAP", colour_by="label")
plotReducedDim(sceFilt, "UMAP", colour_by="cellType")
```

Agreement of clusters with cell types
```{r cluster agreement HC}
tab <- table(sceFilt$label, sceFilt$cellType)
phm <- pheatmap(log10(tab+10), main="cluster vs cell type",
    color=viridis::viridis(100), silent=FALSE)
phm


```

## Result

Sensitivity and specificity of low quality cell detection
```{r sensitivity HC}
mean(sce$discard[sce$cellType == "unassigned"])
mean(!sce$discard[sce$cellType != "unassigned"])

```

Agreement of clustering and cell type

```{r adjusted rand index HC}
mclust::adjustedRandIndex(sceFilt$label,sceFilt$cellType)
```

Runtime in seconds
```{r elapsed HC}
paste(signif((proc.time() - startTime)[ "elapsed"], digits=4), "s")
```